---
title: "Plot the results from the publication M端nch and Laepple (2018)"
author: "Thomas M端nch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting publication results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette documents all the steps needed to obtain the main results
presented in M端nch and Laepple (2018) along with the plotting of the respective
main figures.

## Obtaining the spectral data

```{r, message = FALSE, warning = FALSE}
library(proxysnr)
library(RColorBrewer)
```

Produce the main spectral results for the DML1, DML2 and WAIS oxygen isotope
data sets (i.e. the raw and corrected signal and noise spectra):

```{r}
DWS <- WrapSpectralResults(
    dml1 = dml$dml1, dml2 = dml$dml2, wais = wais,
    diffusion = diffusion.tf,
    time.uncertainty = time.uncertainty.tf,
    df.log = c(0.15, 0.15, 0.1))
```

This function is only a wrapper for the main package functions
`ObtainArraySpectra` and `SeparateSignalFromNoise` which calls the two functions
for all the data sets that are specified as input to `WrapSpectralResults`. The
function `ObtainArraySpectra` is used, for a specific data set, to calculate all
individual spectra, the corresponding mean spectrum and the spectrum of the
stacked record (thus, of the average isotope record in the time domain);
`SeparateSignalFromNoise` is used to obtain the raw and corrected signal and
noise spectra for this data set.

The data sets analysed in the paper are provided along **proxysnr** in the
variables `dml` and `wais`; see `?dml` and `?wais` for details on these data
sets.

The applied transfer functions to correct for the loss in high-frequency
spectral power by the effects of diffusion and time uncertainty are provided in
the variables `diffusion.tf` and `time.uncertainty.tf`, respectively; see
`?diffusion.tf`, `?time.uncertainty.tf`, and the vignette `vignette(topic =
"calculate-transfer-functions", package = "proxysnr")` for details on obtaining
these functions.

The output from `WrapSpectralResults` is a list of the spectral results for each
of the data sets providing the estimated signal, noise and signal-to-noise ratio
spectra (i) without any correction applied ("raw"), (ii) for only applying the
diffusion correction ("corr.diff.only"), (iii) for only applying the time
uncertainty correction ("corr.t.unc.only"), and (iv) for applying both
corrections ("corr.full"):

```{r}
ls.str(DWS)
```

For all or only for some of the data sets, you can omit both or one of the two
transfer functions from the call to `WrapSpectralResults` in which case only the
raw, i.e. uncorrected, or only the partially corrected signal and noise spectra
are returned for these data sets (see `?WrapSpectralResults` for details).

## Plot the main figures

### Plot DML1 isotope array spectra (Figure 1)

```{r, fig.width = 7, fig.height = 6}
PlotArraySpectra(ObtainArraySpectra(dml$dml1, df.log = 0.12),
                 marker = DWS$dml1$corr.full$f.cutoff[2],
		 ylab = expression(
		   "Power spectral density " * "(\u2030"^{2}%.%"yr)"))
```

### Plot DML and WAIS signal and noise spectra (Figure 2)

```{r, warning = FALSE, message = FALSE, fig.width = 14, fig.height = 10.36}
proxysnr:::muench_laepple_fig02(DWS)
```

### Plot frequency dependence of signal-to-noise ratios (Figure 3)

Obtain the final signal-to-noise ratio "spectra" by combining both DML data sets
and applying additional logarithmic smoothing for visual purposes:

```{r}
SNR <- proxysnr:::PublicationSNR(DWS)
```

Plot the figure:

```{r, fig.width = 7, fig.height = 6}
PlotSNR(SNR, f.cut = TRUE, names = c("DML", "WAIS"),
        col = c("black", "dodgerblue4"))
```

### Plot estimated correlation with common signal (Figure 4)

Calculate the estimated correlation of a stacked isotope record with the
underlying common signal as a function of records averaged and the temporal
averaging period (i.e. resolution) of the records:

```{r}
# for the DMl data
crl1 <- ObtainStackCorrelation(SNR$dml, N = 1 : 20,
                               limits = c(1 / 100, SNR$dml$f.cutoff[2]))

# for the WAIS data
crl2 <- ObtainStackCorrelation(SNR$wais, N = 1 : 20,
                               limits = c(1 / 100, SNR$wais$f.cutoff[2]))
```

Plot the figure for DML:

```{r, fig.width = 8, fig.height = 6.3}
PlotStackCorrelation(data = crl1, label = expression(bold("a.")~"DML"),
                     ylim = c(NA, 50))
```

Plot the figure for WAIS:
```{r, fig.width = 8, fig.height = 6.3}
PlotStackCorrelation(data = crl2, label = expression(bold("b.")~"WAIS"),
                     ylim = c(NA, 50))
```

### Plot comparison of DML and Trench noise spectra (Figure 5)

We compare the noise spectrum of the DML firn-core array ("array scale") to the
noise spectrum of the T15 trench oxygen isotope data ("local scale"). The latter
is supplied in the internal package variable `t15.noise` and automatically
loaded by the plotting function:

```{r, fig.width = 7, fig.height = 6}
proxysnr:::muench_laepple_fig05(SNR)
```

## Literature cited

M端nch, T. and Laepple, T.: What climate signal is contained in decadal- to
centennial-scale isotope variations from Antarctic ice cores?, Clim. Past, 14,
2053-2070, doi:
[10.5194/cp-14-2053-2018](https://doi.org/10.5194/cp-14-2053-2018), 2018.

