---
title: "Introduction to proxysnr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to proxysnr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message = FALSE, warning = FALSE}
library(proxysnr)
library(magrittr)
```

## Getting started

With `proxysnr` you apply spectral analyses to climate proxy records that were
obtained from a spatial array of proxy observation locations, e.g. ice cores,
marine sediment cores, tree rings, etc., to derive the power spectral density
(PSD, or, in short, spectrum) of the record's common signal and the PSD of the
local noise (for details see Münch and Laepple, 2018).

To work with `proxysnr`, supply the proxy records in form of a list, or a data
frame, with each list element or data column representing one of the proxy
observation locations, where the proxy data were sampled over time, depth, or
any other given variable, at equal intervals. Note that the equidistance of the
sampling intervals is mandatory to obtain correct spectra.

To showcase the analyses available with `proxysnr` we create an artifical proxy
dataset which mimicks five observation locations, each sampling the proxy data
at 1000 sampling points:

```{r toy-data}
set.seed(20241030)

# five "cores" (proxy observation locations)
nc <- 5
# number of sampling points (e.g. time steps) at each location
nt <- 1000

# simulate a common climate signal as an AR-1 process
clim <- as.numeric(arima.sim(model = list(ar = 0.7), n = nt))
# simulate the independent proxy noise as a white noise process
noise <- as.data.frame(replicate(nc, rnorm(n = nt, sd = 1.5)))

# create a data frame (i.e. a "spatial array") of five "cores" that record the
# same climate signal (scaled to make it relatively a bit stronger) but
# independent noise
cores <- 1.2 * clim + noise
```

## Obtaining and plotting the proxy array spectra

A first step is to calculate the PSD for each proxy record, obtain the average
spectrum of these individual spectra, and get the spectrum of the data series
when averaging across all proxy records (the "stacked" record); all of which is
handled by the function `ObtainArraySpectra()`. We only need to set the sampling
resolution via the parameter `res` (chosen arbitrarily here) in order to obtain
the correct spectral units, and we can apply some smoothing to the raw spectral
estimates (parameter `df.log`) to reduce their spectral uncertainty:

```{r}
arrspec <- ObtainArraySpectra(cores, res = 2, df.log = 0.1)

names(arrspec)
attr(arrspec, "array.par")
```

The output list contains the attribute `array.par` which stores information
on the underlying proxy record array, namely the number of records ("nc"), the
number of sampling points per record ("nt"), and the sampling resolution
("res"), so that other `proxysnr` functions have these information available.

The function `PlotArraySpectra()` directly plots the results:

```{r array-spectra, fig.width = 7, fig.height = 6}
PlotArraySpectra(arrspec, remove = 0, xlim = c(2000, 2), ylim = c(0.05, 50),
                 xlab = "Period (a.u.)", ylab = "Power spectral density")
```

The plot displays the spectra as a function of the inverse spectral frequency,
e.g. time period, which is often more intuitive to interpret. Each proxy
record's spectrum is shown as a thin grey line (however, you can change all
colours to your likings), and the mean spectrum of all individual spectra is
plotted as the thick black line, which averages out some spectral uncertainty of
the individual spectra and thus delivers a more accurate representation of the
average proxy PSD. If all proxy record variations were indepedent noise, the
spectrum of the stacked record (thick brown line) would follow the line of the
mean spectrum divided by the number of records, i.e. the black dashed line on
the plot. Hence, the PSD of the spectrum of the stacked record which exceeds
this line (the blue shaded area) results from proxy variability common to all
records, while the spectral energy which is lost (the red shaded area)
represents independent noise variability.

When you do not need to save the intermediate results, you can also pipe the
workflow,

```{r array-spectra-alt, eval = FALSE}
cores %>%
  ObtainArraySpectra(res = 2, df.log = 0.1) %>%
  PlotArraySpectra(remove = 0, xlim = c(2000, 2), ylim = c(0.05, 50),
                   xlab = "Period (a.u.)", ylab = "Power spectral density")
```

and in the following we always show the combined, piped workflow in order to
highlight how the `proxysnr` functions are linked.

## Signal and noise spectra

While the previous plot is visually insightful, the blue and red shaded areas
are only proportional to the actual signal and noise spectra (see Münch and
Laepple, 2018 for details); these are instead calculated using
`SeparateSignalFromNoise()` which delivers also the spectrum of the
signal-to-noise ratio, which is an important quantity to interpret proxy
variability in terms of climate variability.

```{r sn-spectra}
spec <- cores %>%
  ObtainArraySpectra(res = 2, df.log = 0.1) %>%
  SeparateSignalFromNoise()

str(spec)
```

There are two plotting routines available in `proxysnr`, `LPlot()` and
`LLines()`, to plot any spectral object returned by its functions automatically
on double-logarithmic axes. These follow the concept of base R's `plot()` and
`lines()` functions and can thus make a fresh spectrum plot or add a spectrum to
an existing plot. So to plot the results from our analyses in `spec` we can
write

```{r plot-sn-spectra, fig.width = 7, fig.height = 5.5, fig.show = "hold", out.width = "47%"}
op <- par(mar = c(5, 5, 0.5, 0.5), las = 1,
          cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.25)

LPlot(spec$signal, inverse = TRUE,
      xlab = "Period (a.u.)", ylab = "Power spectral density",
      xlim = c(2000, 2), ylim = c(0.1, 50), col = "dodgerblue4", lwd = 2)
LLines(spec$noise, inverse = TRUE, col = "firebrick4", lwd = 2)
legend("bottomleft", c("Signal", "Noise"), col = c("dodgerblue4", "firebrick4"),
       lty = 1, lwd = 2, bty = "n")

LPlot(spec$snr, inverse = TRUE, xlab = "Period (a.u.)", ylab = "SNR",
      xlim = c(2000, 2), ylim = c(0.1, 50), col = "black", lwd = 2)

par(op)
```

We can test whether the estimated noise spectrum indeed reproduces the standard
deviation of `1.5` of the given white noise process:

```{r noise-sd}
df <- diff(spec$noise$freq)[1]
sqrt(2 * df * sum(spec$noise$spec)) # ok!
```

## Estimating confidence intervals

Since confidence intervals for the signal, noise, and SNR spectra cannot be
obtained analytically, `proxysnr` offers to estimate these from a parametric
bootstrapping procedure based on Monte Carlo simulation of surrogate data:

```{r conf-est}
spec <- cores %>%
  ObtainArraySpectra(res = 2, df.log = 0.1) %>%
  SeparateSignalFromNoise() %>%
  EstimateCI(nmc = 10, df.log = 0.1, ci.df.log = 0.05)
```

The plotting routines `LPlot()` and `LLines()` automatically add a shading for
the confidence intervals to the plot, when such are available, so the same
above plotting code can be used to reproduce the plots now including the only
just estimated confidence intervals:

```{r plot-sn-spectra2, fig.width = 7, fig.height = 5.5, out.width = "47%", echo = FALSE}
op <- par(mar = c(5, 5, 0.5, 0.5), las = 1,
          cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.25)

LPlot(spec$signal, inverse = TRUE,
      xlab = "Period (a.u.)", ylab = "Power spectral density",
      xlim = c(2000, 2), ylim = c(0.1, 50), col = "dodgerblue4", lwd = 2)
LLines(spec$noise, inverse = TRUE, col = "firebrick4", lwd = 2)
legend("bottomleft", c("Signal", "Noise"), col = c("dodgerblue4", "firebrick4"),
       lty = 1, lwd = 2, bty = "n")

LPlot(spec$snr, inverse = TRUE, xlab = "Period (a.u.)", ylab = "SNR",
      xlim = c(2000, 2), ylim = c(0.1, 50), col = "black", lwd = 2)

par(op)
```

## Correction of signal and noise spectra

During archiving, the proxy records may be subject to smoothing, e.g. in firn and
ice by diffusion, or in marine sediments by bioturbation. Furthermore, if the
proxy records are dated their chronologies may be uncertain between the
records, with this time uncertainty affecting the reconstruction of the common
signal.

`proxysnr` offers to correct the estimated signal and noise spectra for these
effects by supplying respective transfer functions to
`SeparateSignalFromNoise()` via the function parameters `diffusion` and
`time.uncertainty`. These transfer functions describe the loss in spectral power
due to the smoothing or the uncertain chronologies and need to be assessed and
provided by the user.

For two special cases you can apply `proxysnr` to directly calculate the
transfer functions: firn diffusion (function `CalculateDiffusionTF()`) and time
uncertainty in _layer-counted_ chronologies (function
`CalculateTimeUncertaintyTF()`); see also Münch and Laepple (2018) for more
details on the processes and the implemented estimation, and the
`vignette("calculate-transfer-functions")` for details on using the two
functions.

Proxy records also contain measurement noise. If you have an estimate of its
magnitude and its spectral shape, you can supply these information to
`SeparateSignalFromNoise()` via the `measurement.noise` parameter, and the
returned noise and SNR spectra will be corrected for the measurement noise,
i.e. they will take into account only the proxy-specific noise that was created
during the archiving process.

## Further analysis options

There are some more `proxysnr` functions you can use to conduct and visualise
analyses:

- `GetIntegratedSNR()` delivers the ratio of the integrated signal and noise
  spectra, which corresponds to the SNR at a specific sampling resolution;
- `ObtainStackCorrelation()` calculates the expected correlation of a stack of
  proxy records with the common climate signal as a function of the number of
  records averaged and the records' sampling resolution (see Fig. 4 in Münch and
  Laepple, 2018 for an example);
- `PlotStackCorrelation()` to directly plot the output from
  `ObtainStackCorrelation()`;
- `PlotSNR()` to plot several SNR datasets in one figure;
- `PlotTF()` to make a plot of the transfer functions obtained with
  `CalculateDiffusionTF()` and/or `CalculateTimeUncertaintyTF()`.

## Package data

`proxysnr` ships with four package datasets:

- `?dml` contains the oxygen isotope data of 15 firn cores from Dronning Maud
  Land (DML), Antarctica (Graf et al., 2002);
- `?wais` contains the oxygen isotope data of 5 firn cores from the West
  Antarctic Ice Sheet (WAIS; Steig, 2013);
- `?diffusion.tf` contains pre-calculated diffusion transfer functions for the
  DML and WAIS isotope records; and
- `?time.uncertainty.tf` contains pre-calculated time uncertainty transfer
  functions for these records.

The datasets can be used to test out the functions from `proxysnr`, and they
have been the basis for the analyses in Münch and Laepple (2018); see also the
`vignette("plot-muench-laepple-figures")` for reproducing their results.

## References

Graf, W., et al.: Stable-isotope records from Dronning Maud Land, Antarctica,
PANGAEA, doi: [10.1594/PANGAEA.728240](https://doi.org/10.1594/PANGAEA.728240),
2002.

Münch, T. and Laepple, T.: What climate signal is contained in decadal- to
centennial-scale isotope variations from Antarctic ice cores?, Clim. Past, 14,
2053-2070, doi:
[10.5194/cp-14-2053-2018](https://doi.org/10.5194/cp-14-2053-2018), 2018.

Steig, E. J.: West Antarctica Ice Core and Climate Model Data, U.S. Antarctic
Program (USAP) Data Center, doi:
[10.7265/N5QJ7F8B](https://doi.org/10.7265/N5QJ7F8B), 2013.
